---
title: "R Notebook"
output: html_notebook
summary: the script obtains the list of targeted therapy drugs from Bioportal's NCIT ontology 
TODO: add checking the drugs and groups files from prev. download and calculating diffs, keeping the established IDs intact.
---

```{r include=FALSE}
library(curl)
library(httr)
library(purrr)
library(magrittr)
library(jsonlite)
library(stringr)
library(openssl)
```

Provide your own API key here
```{r}
api_key = ""
targ_ther_drugs_tbl_name <- "dict.targeted_therapy_drugs"
targ_ther_groups_tbl_name <- "dict.targeted_therapy_groups"
```

Fetching Bioportal's NCIT to retrieve targeted therapy graphs
```{r}
targ_ther <- GET(paste0("https://data.bioontology.org/ontologies/NCIT/classes/http%3A%2F%2Fncicb.nci.nih.gov%2Fxml%2Fowl%2FEVS%2FThesaurus.owl%23C163758/children?apikey=", 
                        api_key, 
                        collapse = "")
                 )

if (targ_ther$status_code == 200) {
        targ_ther <- targ_ther$content %>%
        rawToChar() %>% 
        parse_json()
} else {
    cat("Error during fetching Bioportal. Code:", targ_ther$status_code)
}
```


Retrieving drugs' paths traversing the graphs
```{r}
process_child_url <- function(url) {
    cat("Processing URL:", url)
    
    resp <- GET(paste0(url, "?apikey=", api_key,
                       collapse = "")
                )
    
    cat("  DONE", "\n")
    
    return(resp)
}

retrieve_children <- function(node, api_key, parent_node_preflabel) {
    if (node$totalCount == 0) {
        return(parent_node_preflabel)   
    }
    
    node$collection <- node$collection
    
    # Children names
    node_collection_preflabels <- node$collection %>% 
        map_chr(., ~ .x$prefLabel) 
    # Term's synonyms
    node_collection_synonyms <- map_chr(node$collection, 
                                        ~ paste0(.x$synonym, collapse = ";"))
    node_collection_preflabels <- map2_chr(node_collection_preflabels,
                                           node_collection_synonyms,
                                           ~ paste(.x, .y, sep = ";"))
    
    # Adding children names to paths
    node_hierarchy <- map_chr(node_collection_preflabels,
                              ~ paste(parent_node_preflabel, .x, sep = "|")
                              )
    
    # Fetching children links
    children_urls <- node$collection %>% map_chr(., ~ .x$links$children)
    
    children <- map(children_urls, process_child_url)
    
    children <- map(children, ~ .x$content %>%
                        rawToChar() %>%
                        parse_json())

    children %<>% map2(., node_hierarchy, 
                       ~ retrieve_children(.x, api_key, .y)
                       )
}

root_preflabel <- "Targeted Therapy Agent"

drug_paths <- retrieve_children(targ_ther, api_key, root_preflabel)
drug_paths %<>% unlist(., recursive = T)
```


```{r}
drug_graphs <- list()

for (i in seq_along(drug_paths)) {
    node = strsplit(drug_paths[i], "\\|")
    subnodes = strsplit(node[[1]], "\\;")
    
    for (j in seq_along(subnodes)) {
        names(subnodes)[j] = subnodes[1]
    }
    
    drug_graphs[[i]] = subnodes
}
```


```{r}
remove_names_duplicates <- function(node) {
    group_names <- node[1:length(node) - 1] %>% 
        unlist(use.names = F) %>% 
        tolower() %>% 
        unique() %>% 
        extract(1:2)
    
    group_names %<>% extract(
        group_names %>%
            is.na() %>% 
            not() %>% 
            which()
        )
    
    drug_names <- node[[length(node)]] %>% 
        tolower() %>% 
        unique() %>% 
        extract(1:3)
    
    drug_names %<>% extract(
        drug_names %>%
            is.na() %>% 
            not() %>% 
            which()
        )
    
    drug_names %<>% str_replace_all("'", '"')
    
    out <- list(group_names, drug_names)
    
    return(out)
}

drugs_graphs %<>% map(., function(x) { 
    x[[1]] <- NULL
    return(x)
    })

drugs_graphs %<>% map(remove_names_duplicates)
```

Retrieve drugs and their groups producing two sep. lists
```{r}
drugs <- drugs_graphs %>% map(., function(x) { 
    if (length(x[[length(x)]]) < 2) {
        return(x[[length(x)]])
    } else {
        return(x[[length(x)]][1:2])
    }
    })

extract_groups <- function (node) {
    if (length(node) == 2) {
        out <- "no group"
    } else if (length(node) == 3) {
        out <- node[[2]][1] %>% tolower() %>% unique()
    } else {
        out <- map_chr(2:(length(node) - 1), ~ node[[.x]][1] %>% 
                           tolower() %>% 
                           unique())
    }
    
    return(out)
}

groups <- tree %>% map(., extract_groups)
```

Groups within the targeted therapy class are of hierarchical structure. The group 
hierarchy graphs are constructed here.
```{r}
groups_hierarchy <- list()

retrieve_groups <- function(group_node) {
    if (group_node[1] == "no groups") return()
    
    i = 1
    while(!is.na(group_node[i])) {
        row_num <- (names(groups_hierarchy) %in% group_node[i]) %>% 
            which()
        
        if (!is_empty(row_num)) {
            if (!is.na(group_node[i + 1])) { 
                # Existing children
                chd             <- groups_hierarchy[[row_num]][[3]]
                chd_position    <- match(group_node[i + 1], chd)
                                
                # If the current group's child is new, append it
                if (is.na(chd_position)) {
                     chd %<>% append(group_node[i + 1])
                     groups_hierarchy[[row_num]][[3]] <<- chd
                }
                
                # Searching the child within the already added nodes 
                chd_index = match(group_node[i + 1], names(groups_hierarchy))
                
                # Setting the node as its child's parent 
                if (!is.na(chd_index)) {
                    groups_hierarchy[[chd_index]][[2]] <<- 
                        append(groups_hierarchy[[chd_index]][[2]], group_node[i])
                } else {
                    groups_hierarchy[[length(groups_hierarchy) + 1]] <<-
                        vector("list", 3L)
                    new_len                             <- length(groups_hierarchy)
                    groups_hierarchy[[new_len]][[1]]    <<- group_node[i + 1]
                    names(groups_hierarchy)[new_len]    <<- group_node[i + 1]
                    groups_hierarchy[[new_len]][[2]]    <<- group_node[i]
                }
            }
        
            i = i + 1
            next
        }  
        
        cat(group_node[i], "new\n")
        groups_hierarchy[[length(groups_hierarchy) + 1]]    <<- vector("list", 3L)
        new_len                                             <- length(groups_hierarchy)
        groups_hierarchy[[new_len]][[1]]                    <<- group_node[i]
        names(groups_hierarchy)[new_len]                    <<- group_node[i]
        
        # If the node has children, add them
        if (!is.na(group_node[i + 1])) { 
            groups_hierarchy[[length(groups_hierarchy)]][[3]] <<- 
                group_node[i + 1]
            
            chd_index = match(group_node[i + 1], names(groups_hierarchy))
                                
            # Setting the node as its child's parent 
            if (!is.na(chd_index)) {
                groups_hierarchy[[chd_index]][[2]] <<- 
                    append(groups_hierarchy[[chd_index]][[2]], group_node[i])
            } else {
                groups_hierarchy[[length(groups_hierarchy) + 1]] <<-
                    vector("list", 3L)
                
                new_len = length(groups_hierarchy)
                groups_hierarchy[[new_len]][[1]]    <<- group_node[i + 1]
                names(groups_hierarchy)[new_len]    <<- group_node[i + 1]
                groups_hierarchy[[new_len]][[2]]    <<- group_node[i]
            }
        }
        
        i = i + 1
    }
}

out <- map(groups, retrieve_groups)
rm(out)

# Cleaning node's parents and children removing duplicates
groups_hierarchy %<>% map(., ~ map(.x, ~ .x %<>% unique()))

groups_hierarchy[[groups_hierarchy %>% 
                     map_lgl(~ .x[[1]] == 'no group') %>% 
                     which()]] <- NULL
```

Producing SQL queries for the groups table
```{r}
# YOUR CURRENT MAX CONCEPT'S ID (FROM THE dict.concepts table) MUST BE HERE. 
# PLEASE, REPLACE THE VALUE OR YOU MAY GET ERRORS WHEN EXECUTING THE QUERY LATER
start_id <- 4270938 # current max. concept's ID + 1    
names(groups_hierarchy) = start_id:(start_id + length(groups_hierarchy) - 1)

produce_groups_sql_query <- function(node, id) {
    out <- list()
    grp_name <- paste0("'", node[[1]], "'", collapse = "")
    out <- out %>% append(id) %>% append(grp_name)
    
    i = 1
    for (i in 2:length(node)) {
        if (!is_empty(node[[i]])) {
            arr <- node[[i]] %>% 
                paste0("'", ., "'", collapse = ", ") %>% 
                paste0("ARRAY [", ., "]", collapse = " ")
        } else {
            arr <- node[[i]] %>% 
                paste0("ARRAY [", ., "]::varchar[]", collapse = " ")
        }
        out %<>% append(arr)
    }
    
    out <- out %>% 
        paste0(., collapse = ", ") %>% 
        paste0("(", ., ")", collapse = " ")
    
    return(out)
}

groups_sql_queries <- map2(groups_hierarchy, 
                           start_id:(start_id + length(groups_hierarchy) - 1),
                           produce_groups_sql_query)
groups_sql_queries %<>% paste0(collapse = ",\n")
groups_sql_queries %<>% paste0("insert into ", targ_ther_groups_tbl_name, " values\n", .,
                               collapse = " ")
```

Writing groups SQL queries to file
```{r}
curr_time <- format(Sys.time(), "%d-%b-%Y-%H:%M")

write(groups_sql_queries, 
      file = paste0("/home/user/Data/targ_ther_groups_", 
                    curr_time, ".sql", collapse = "")
      )
```


Matching drugs' groups with their IDs
```{r}
unique_groups <- groups %>% 
    unlist() %>%
    unique()

group_names <- unique_groups
group_ids   <- start_id:(start_id + length(unique_groups) - 1)
group_mapping  <- data.frame(id = group_ids,
                             name = group_names)

assign_group_id <- function(grp_node, groups_ids) {
    grp_node <- map_int(grp_node, ~ groups_ids[match(.x, groups_ids[, 2]), 1])
    return(grp_node)
}

groups %<>% map(., ~ assign_group_id(.x, groups_ids))
```

Removing duplicates from drugs
```{r}
duplicates <- vector("logical", length(drugs))

for (i in seq_along(drugs)) {
    if (i < 2) next
    
    drug_index = -1
    
    for (j in seq_along(drugs[1:(i - 1)])) {
        if (drugs[[j]][1] == drugs[[i]][1]) {
            drug_index = j
            break
        }
    }
    group_index <- groups[[i]][1] %in% groups[drug_index]
    
    if (drug_index != -1 && isTRUE(group_index)) {
        duplicates[i] <- TRUE
    } else if (drug_index != -1 && !isTRUE(group_index)) {
        groups[[drug_index]] %<>% append(., groups[[i]])
        groups[[drug_index]] %<>% unique()
        duplicates[i] <- TRUE
    }
}

drugs %<>% .[duplicates %>% not() %>% which()]
groups %<>% .[duplicates %>% not() %>% which()]
```


```{r}
produce_drugs_sql_query <- function(drug_node, group_node) {
    drug_arr <- drug_node %>% 
        paste("'", ., "'",
              sep = "", collapse = ", ") %>% 
        paste0("ARRAY [", ., "]", collapse = "")
    
    if (is.na(group_node[1])) {
        group_arr <- "ARRAY []::integer[]"
    } else {
        group_arr <- group_node %>% 
        paste0(collapse = ", ") %>% 
        paste0("ARRAY [", ., "]", collapse = "")   
    }
    
    out <- paste0("(", drug_arr, ", ", group_arr, ")",
                  collapse = " ")
}

targ_ther_drugs_sql <- map2(drugs, groups, produce_drugs_sql_query)
targ_ther_drugs_sql <- targ_ther_drugs_sql %>% 
    paste0(collapse = ",\n") %>% 
    paste0("insert into", targ_ther_drugs_tbl_name, "(drug_aliases, drug_groups) values\n", .,
           collapse = "")
```


```{r}
curr_time <- format(Sys.time(), "%d-%b-%Y-%H:%M")

write(targ_ther_drugs_sql, 
      file = paste0("/home/user/Data/targ_ther_drugs_", 
                    curr_time, ".sql", collapse = "")
      )
```

